---
title: "Data Simulation with sccomp_simulate"
author: "sccomp"
date: today
format:
  html:
    toc: true
    code-fold: true
    code-tools: true
---

## Introduction

The `sccomp_simulate` function allows you to simulate compositional count data from a fitted sccomp model. This vignette demonstrates three main use cases:

1. **Posterior Predictive Checks**: Simulate data using the full fitted model (all parameters from posterior distribution)
2. **New Data Simulation**: Simulate data for new sample conditions using the fitted model
3. **Custom Coefficients**: Simulate data with user-specified coefficients for specific scenarios

## Setup

```{r setup, message=FALSE, warning=FALSE}
# Load development version to ensure latest features
if(requireNamespace("devtools", quietly = TRUE)) {
  devtools::load_all(quiet = TRUE)
}
library(sccomp)
library(dplyr)
library(ggplot2)
library(tidyr)
```

Load example data:

```{r load-data, message=FALSE}
data("counts_obj")
```

## 1. Posterior Predictive Check (Using Everything from Fit)

The simplest use case is to simulate data using all parameters from the fitted model. This is useful for posterior predictive checks to validate model fit.

### Fit the Model

```{r fit-model, eval = instantiate::stan_cmdstan_exists(), message=FALSE}
estimate = sccomp_estimate(
  counts_obj,
  formula_composition = ~ type, 
  formula_variability = ~ 1,
  sample = "sample", 
  cell_group = "cell_group", 
  abundance = "count",
  
  verbose = FALSE,
  max_sampling_iterations = 2000
)
```

### Simulate from Posterior Distribution

When `new_data = NULL` and `coefficients = NULL`, the function uses the original data and samples from the posterior distribution of all model parameters:

```{r posterior-predictive, eval = instantiate::stan_cmdstan_exists(), message=FALSE}
# Simulate using all parameters from the fitted model
simulated_data = sccomp_simulate(
  estimate,
  formula_composition = ~ type,
  formula_variability = ~ 1,
  new_data = NULL,        # Use original data
  coefficients = NULL,    # Use posterior beta_raw
  
  number_of_draws = 10     # Generate 10 replicates
)

# View the simulated data
head(simulated_data, 20)
```

The simulated data includes:
- Original columns from the input data
- `generated_counts`: Simulated cell counts
- `replicate`: Which draw from the posterior (1, 2, 3, etc.)

### Visualize Posterior Predictive Check

```{r visualize-ppc, eval = instantiate::stan_cmdstan_exists(), message=FALSE, fig.height=6}
# Compare observed vs simulated
comparison = 
  bind_rows(
    counts_obj |> 
      mutate(type = "observed") |>
      select(sample, cell_group, count, type),
    simulated_data |>
      filter(replicate == 1) |>
      mutate(type = "simulated") |>
      select(sample, cell_group, generated_counts, type) |>
      rename(count = generated_counts)
  )

comparison |>
  ggplot(aes(x = cell_group, y = count, fill = type)) +
  geom_boxplot(position = "dodge", alpha = 0.7, outlier.alpha = 0.3) +
  scale_y_continuous(trans = "log1p") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Posterior Predictive Check: Observed vs Simulated",
       subtitle = "All samples combined",
       x = "Cell Group", y = "Count (log1p scale)")
```

## 2. Simulate for New Data (New Sample Conditions)

You can simulate data for new sample conditions (e.g., new treatment groups, new time points) while using the fitted model parameters.

### Create New Data

When `coefficients = NULL`, `new_data` must include the `cell_group` column. We can expand the sample data to include all cell groups:

```{r new-data-example, eval = instantiate::stan_cmdstan_exists(), message=FALSE}
# Create new sample conditions
sample_conditions = tibble(
  sample = c("new_sample_1", "new_sample_2", "new_sample_3"),
  type = factor(c("normal", "cancer", "cancer"))
)

# Expand to include all cell groups from the fitted model
# Get unique cell groups from original data
all_cell_groups = counts_obj |>
  distinct(cell_group) |>
  pull(cell_group)

# Create full new_data with all combinations
new_samples = sample_conditions |>
  tidyr::crossing(cell_group = all_cell_groups)

# Simulate for new samples using posterior parameters
simulated_new = sccomp_simulate(
  estimate,
  formula_composition = ~ type,
  formula_variability = ~ 1,
  new_data = new_samples,
  coefficients = NULL,    # Still use posterior beta_raw
  
  number_of_draws = 10
)

head(simulated_new, 20)
```

Note: When `coefficients = NULL`, the function automatically expands `new_data` to include all cell groups from the fitted model. The design matrix for the new samples is created based on the `formula_composition` and `formula_variability`.

### Simulate for Subset of Samples

You can also simulate for a subset of the original samples. When `coefficients = NULL`, the data must include `cell_group`:

```{r subset-samples, eval = instantiate::stan_cmdstan_exists(), message=FALSE}
# Select a subset of original samples (with cell_group)
subset_samples = counts_obj |>
  filter(sample %in% unique(counts_obj$sample)[1:3]) |>  # First 3 samples
  select(sample, type, cell_group)

# Simulate for subset
simulated_subset = sccomp_simulate(
  estimate,
  formula_composition = ~ type,
  formula_variability = ~ 1,
  new_data = subset_samples,
  coefficients = NULL,
  cores = 1
)

head(simulated_subset, 15)
```

## 3. Simulate with Custom Coefficients

You can specify custom coefficients to simulate specific scenarios (e.g., hypothetical treatment effects, what-if analyses).

### Create Coefficients Table

The coefficients table must contain:
- A column matching the `cell_group` column name
- Columns matching the design matrix column names (e.g., `(Intercept)`, `typecancer`)

```{r create-coefficients, eval = instantiate::stan_cmdstan_exists(), message=FALSE}
# Get design matrix column names from the fitted model
# These typically include "(Intercept)" and factor levels
design_cols = c("(Intercept)", "typecancer")

# Create coefficients table
# Example: Set all coefficients to zero (null scenario)
coeffs_null = counts_obj |>
  distinct(cell_group) |>
  mutate(
    `(Intercept)` = 0,
    `typecancer` = 0
  )

coeffs_null
```

### Verify Zero Coefficients Give Equal Mean Proportions

When all coefficients are zero (both intercept and treatment effects), all cell types should have equal mean proportions (1/M, where M is the number of cell types). This is because with zero log-odds, `softmax(0, 0, ..., 0) = (1/M, 1/M, ..., 1/M)`.

Let's verify this by simulating 30 samples:

```{r verify-zero-coefficients, eval = instantiate::stan_cmdstan_exists(), message=FALSE, fig.height=6}
# Simulate with zero coefficients across 30 samples
sample_data_many = tibble(
  sample = paste0("sample_", 1:30),
  type = factor(rep(c("normal", "cancer"), each = 15))
)

sim_zero = sccomp_simulate(
  estimate,
  formula_composition = ~ type,
  formula_variability = ~ 1,
  new_data = sample_data_many,
  coefficients = coeffs_null,
  
  number_of_draws = 10
)

# Expected proportion (1/M where M is number of cell types)
M = nrow(coeffs_null)
expected_prop = 1 / M

cat("Expected proportion per cell type (1/M):", round(expected_prop, 4), "\n")
cat("Number of cell types (M):", M, "\n\n")

# Visualize with boxplots - all cell types should have similar mean proportions
sim_zero |>
  ggplot(aes(x = cell_group, y = generated_proportions)) +
  geom_boxplot(fill = "steelblue", alpha = 0.7, outlier.alpha = 0.3) +
  geom_hline(
    yintercept = expected_prop, 
    linetype = "dashed", 
    color = "red", 
    linewidth = 1
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(
    title = "Proportions with Zero Coefficients (30 samples)",
    subtitle = paste("Expected mean:", round(expected_prop, 4), "(red dashed line) - All cell types should align"),
    x = "Cell Group",
    y = "Proportion",
    caption = "With all coefficients = 0, all cell types should have equal mean proportions (1/M)"
  )
```

**Note**: Individual samples may show variation due to:
- Random effects (if present in the model)
- Variability parameters (alpha) affecting the beta-binomial distribution
- Sampling variability

However, the **mean** across many samples should converge to 1/M for each cell type.

### Simulate with Custom Coefficients

```{r simulate-custom-coefficients, eval = instantiate::stan_cmdstan_exists(), message=FALSE}
# Create new data for samples
new_samples_coef = tibble(
  sample = c("scenario_1", "scenario_2"),
  type = factor(c("normal", "cancer"))
)

# Simulate with custom coefficients
simulated_custom = sccomp_simulate(
  estimate,
  formula_composition = ~ type,
  formula_variability = ~ 1,
  new_data = new_samples_coef,
  coefficients = coeffs_null,
  
  number_of_draws = 10
)

head(simulated_custom, 20)

# Note: With zero coefficients, mean proportions should be equal across cell types
# Individual samples may vary due to random effects and variability parameters,
# but the mean across many samples should converge to 1/M for each cell type
```

### Simulate Different Scenarios

You can create multiple coefficient tables to simulate different scenarios. When using `new_data`, it's useful to simulate from at least 30 samples to see clear patterns:

```{r multiple-scenarios, eval = instantiate::stan_cmdstan_exists(), message=FALSE}
# Create sample data with 30 samples for better visualization
new_samples_scenarios = tibble(
  sample = paste0("scenario_sample_", 1:30),
  type = factor(rep(c("normal", "cancer"), each = 15))
)

# Scenario 1: Strong cancer effect for specific cell type
coeffs_strong_effect = counts_obj |>
  distinct(cell_group) |>
  mutate(
    `(Intercept)` = 0,
    `typecancer` = if_else(cell_group == "B1", 2.0, 0.0)  # Strong effect for B1
  )

# Scenario 2: Moderate effect across all cell types
coeffs_moderate = counts_obj |>
  distinct(cell_group) |>
  mutate(
    `(Intercept)` = 0,
    `typecancer` = 0.5  # Moderate effect for all
  )

# Simulate both scenarios
scenario1 = sccomp_simulate(
  estimate,
  formula_composition = ~ type,
  formula_variability = ~ 1,
  new_data = new_samples_scenarios,
  coefficients = coeffs_strong_effect,
  cores = 1
) |> mutate(scenario = "Strong B1 effect")

scenario2 = sccomp_simulate(
  estimate,
  formula_composition = ~ type,
  formula_variability = ~ 1,
  new_data = new_samples_scenarios,
  coefficients = coeffs_moderate,
  cores = 1
) |> mutate(scenario = "Moderate effect")

# Compare scenarios - focus on cancer samples
bind_rows(scenario1, scenario2) |>
  filter(type == "cancer") |>
  ggplot(aes(x = cell_group, y = generated_counts, fill = scenario)) +
  geom_boxplot(position = "dodge", alpha = 0.7) +
  scale_y_continuous(trans = "log1p") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Comparison of Different Coefficient Scenarios (30 cancer samples)",
       x = "Cell Group", y = "Simulated Counts (log1p scale)")
```

## 4. Combining New Data with Custom Coefficients

When both `new_data` and `coefficients` are provided, `new_data` can omit the `cell_group` column. The function automatically expands `new_data` to include all cell groups from the `coefficients` table using a cross join:

```{r combine-newdata-coefficients, eval = instantiate::stan_cmdstan_exists(), message=FALSE}
# Sample-specific data (no cell_group needed)
sample_data = tibble(
  sample = c("experiment_1", "experiment_2", "experiment_3"),
  type = factor(c("normal", "cancer", "cancer"))
)

# Cell-type-specific coefficients
coeffs_table = counts_obj |>
  distinct(cell_group) |>
  mutate(
    `(Intercept)` = 0,
    `typecancer` = 0
  )

# Simulate: new_data specifies samples, coefficients specifies cell types
simulated_combined = sccomp_simulate(
  estimate,
  formula_composition = ~ type,
  formula_variability = ~ 1,
  new_data = sample_data,      # Sample-specific
  coefficients = coeffs_table,  # Cell-type-specific
  cores = 1
)

# The result has all combinations of samples x cell_groups
simulated_combined |>
  select(sample, type, cell_group, generated_counts) |>
  head(20)
```

## 5. Advanced: Multiple Draws and Variability

### Custom Mean-Dispersion Association Slope

The `mean_dispersion_slope` parameter allows you to override the slope of the mean-dispersion association from the fitted model. This controls how variability (dispersion) relates to mean abundance. If `NULL`, the slope from the fitted model (`prec_coeff[2]`) is used.

We can visualize the effect by comparing 2D plots from `plot_2D_intervals()` showing the mean-variance association:

```{r mean-dispersion-slope, eval = instantiate::stan_cmdstan_exists(), message=FALSE, fig.height=8}
# First, get the 2D plot from the original fit (shows the association)
# The original estimate was fitted with formula_variability = ~ 1, but we need ~ type for 2D plots
# So we'll refit with ~ type for variability
estimate_with_variability = sccomp_estimate(
  counts_obj,
  formula_composition = ~ type,
  formula_variability = ~ type,  # Include type in variability to get 2D plots for factor
  sample = "sample",
  cell_group = "cell_group",
  abundance = "count",
  verbose = FALSE
)

estimate_tested = estimate_with_variability |>
  sccomp_test()

plot_original = plot_2D_intervals(estimate_tested) +
  labs(title = "Original Fit: Mean-Dispersion Association")

# Simulate data with zero slope (no association)
# When mean_dispersion_slope is provided, the intercept is automatically preserved from the original fit
# Use formula_variability = ~ type to get 2D plots for the factor as well
# Use multiple replicates and treat them as separate samples for better fit
sim_no_association = sccomp_simulate(
  estimate_with_variability,  # Use the refitted estimate with ~ type variability
  formula_composition = ~ type,
  formula_variability = ~ type,  # Include type in variability to get 2D plots for factor
  new_data = NULL,
  coefficients = NULL,
  mean_dispersion_slope = 0.0,  # No association (intercept is automatically preserved)
  
  number_of_draws = 10  # Use multiple replicates
)

# Get the original fit parameters for comparison
fit_original = attr(estimate_tested, "fit")
prec_coeff_original = fit_original$summary("prec_coeff")
cat("Original fit: intercept =", round(prec_coeff_original$mean[1], 2), 
    ", slope =", round(prec_coeff_original$mean[2], 2), "\n")
cat("Simulation used: intercept =", round(prec_coeff_original$mean[1], 2), 
    ", slope = 0.0\n")

# Fit the simulated data (with zero association)
# Treat each replicate as a separate sample to get more data
sim_data_for_fit = sim_no_association |>
  mutate(sample = paste0(sample, "_rep", replicate)) |>
  select(sample, cell_group, type, count = generated_counts)

estimate_simulated = sccomp_estimate(
  sim_data_for_fit,
  formula_composition = ~ type,
  formula_variability = ~ type,  # Match the variability formula used in simulation
  sample = "sample",
  cell_group = "cell_group",
  abundance = "count",
  
  verbose = FALSE
) |>
  sccomp_test()

# Get the fitted parameters from simulated data
fit_simulated = attr(estimate_simulated, "fit")
prec_coeff_simulated = fit_simulated$summary("prec_coeff")
cat("Fitted from simulated data: intercept =", round(prec_coeff_simulated$mean[1], 2), 
    ", slope =", round(prec_coeff_simulated$mean[2], 2), "\n")

plot_simulated = plot_2D_intervals(estimate_simulated) +
  labs(title = "Simulated with Slope = 0: No Mean-Dispersion Association")

# Display plots side by side
library(patchwork)
plot_original / plot_simulated

# Note: The regression line in the plot shows the ESTIMATED parameters from fitting the simulated data,
# not the simulation parameters themselves. The simulation correctly used:
# - intercept = original fit intercept (preserved automatically)
# - slope = 0.0 (as specified)
# However, when we fit the simulated data, the model estimates new parameters from the data structure.
# The fitted intercept may differ from the simulation intercept because:
# - The simulated data has a different structure (slope = 0 changes the alpha-beta relationship)
# - The model estimates parameters that best fit the observed data
# The "(Intercept, adjusted)" facet shows the variability effect after adjusting for
# the estimated association, which should be closer to horizontal when slope = 0.
```

### Simulate with Non-Zero Coefficients and No Association

We can also simulate with specific coefficients (e.g., a treatment effect) while removing the mean-dispersion association:

```{r simulate-with-coefficients-no-association, eval = instantiate::stan_cmdstan_exists(), message=FALSE, fig.height=10}
# Create coefficients table with a non-zero effect for typecancer
# We'll use a moderate effect size
# Coefficients must sum to zero for compositional models
coeffs_no_assoc = estimate_tested |>
  distinct(cell_group) |>
  mutate(
    `(Intercept)` = 0,
    typecancer_raw = if_else(cell_group == "B1", 1.5, 0)  # B1 has positive effect
  ) |>
  mutate(
    typecancer = typecancer_raw - mean(typecancer_raw)  # Normalize to sum to zero
  ) |>
  select(-typecancer_raw)

# Simulate with these coefficients and slope = 0 (no association)
# Use formula_variability = ~ type to get 2D plots for the factor as well
sim_coeffs_no_assoc = sccomp_simulate(
  estimate_with_variability,  # Use the refitted estimate with ~ type variability
  formula_composition = ~ type,
  formula_variability = ~ type,  # Include type in variability to get 2D plots for factor
  new_data = NULL,
  coefficients = coeffs_no_assoc,
  mean_dispersion_slope = 0.0,  # No association
  
  number_of_draws = 10
)

cat("Simulation with coefficients and no association completed\n")
cat("Coefficients used: B1 typecancer =", coeffs_no_assoc$typecancer[coeffs_no_assoc$cell_group == "B1"], "\n")

# Fit the simulated data
sim_data_coeffs = sim_coeffs_no_assoc |>
  mutate(sample = paste0(sample, "_rep", replicate)) |>
  select(sample, cell_group, type, count = generated_counts)

estimate_simulated_coeffs = sccomp_estimate(
  sim_data_coeffs,
  formula_composition = ~ type,
  formula_variability = ~ type,  # Match the variability formula used in simulation
  sample = "sample",
  cell_group = "cell_group",
  abundance = "count",
  
  verbose = FALSE
) |>
  sccomp_test()

# Get fitted parameters
fit_simulated_coeffs = attr(estimate_simulated_coeffs, "fit")
prec_coeff_simulated_coeffs = fit_simulated_coeffs$summary("prec_coeff")
cat("Fitted from simulated data (with coefficients): intercept =", round(prec_coeff_simulated_coeffs$mean[1], 2), 
    ", slope =", round(prec_coeff_simulated_coeffs$mean[2], 2), "\n")

plot_simulated_coeffs = plot_2D_intervals(estimate_simulated_coeffs) +
  labs(title = "Simulated with Coefficients (b1≠0) and Slope = 0: No Mean-Dispersion Association")

# Display all three plots: original, simulated without coefficients, simulated with coefficients
plot_original / plot_simulated / plot_simulated_coeffs
```

The three plots show:
1. **Top**: Original fit with mean-dispersion association (sloped regression line)
2. **Middle**: Simulated with zero coefficients and slope = 0 (no association, horizontal line)
3. **Bottom**: Simulated with non-zero coefficients (B1 has positive effect) and slope = 0 (no association, horizontal line)

Even with non-zero coefficients, when `mean_dispersion_slope = 0`, there should be no association between mean and dispersion, resulting in a horizontal regression line.
```

### Multiple Posterior Draws

You can simulate multiple draws from the posterior distribution:

```{r multiple-draws, eval = instantiate::stan_cmdstan_exists(), message=FALSE}
# Simulate 5 draws from posterior
simulated_multi = sccomp_simulate(
  estimate,
  formula_composition = ~ type,
  formula_variability = ~ 1,
  new_data = NULL,
  coefficients = NULL,
  
  number_of_draws = 10
)

# Check replicate column
simulated_multi |>
  distinct(replicate) |>
  count()
```

### Variability Multiplier

You can artificially increase variability for benchmarking or sensitivity analysis:

```{r variability-multiplier, eval = instantiate::stan_cmdstan_exists(), message=FALSE}
# Standard simulation
sim_normal = sccomp_simulate(
  estimate,
  formula_composition = ~ type,
  formula_variability = ~ 1,
  new_data = NULL,
  coefficients = NULL,
  
  variability_multiplier = 1  # Default
) |> mutate(variability = "Normal")

# Increased variability
sim_high = sccomp_simulate(
  estimate,
  formula_composition = ~ type,
  formula_variability = ~ 1,
  new_data = NULL,
  coefficients = NULL,
  
  variability_multiplier = 10  # 10x variability
) |> mutate(variability = "High")

# Compare
bind_rows(sim_normal, sim_high) |>
  filter(replicate == 1) |>
  ggplot(aes(x = cell_group, y = generated_counts, fill = variability)) +
  geom_boxplot(position = "dodge", alpha = 0.7) +
  scale_y_continuous(trans = "log1p") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  labs(title = "Effect of Variability Multiplier",
       x = "Cell Group", y = "Simulated Counts (log1p scale)")
```

## 6. Comparing Simulated Scenarios: Significant Cell Type Analysis

A powerful use case is comparing simulations with different coefficient scenarios, especially for cell types that show significant differences. This helps visualize the impact of treatment effects vs. null scenarios.

### Find a Significantly Different Cell Type

First, let's identify which cell types are significantly different between conditions:

```{r find-significant, eval = instantiate::stan_cmdstan_exists(), message=FALSE}
# Test for significance
test_results = estimate |>
  sccomp_test()

# Find significantly different cell types (e.g., FDR < 0.05)
significant_cells = test_results |>
  filter(c_FDR < 0.05 | is.na(c_FDR)) |>
  arrange(c_FDR) |>
  head(3)

# Display significant cell types
significant_cells |>
  select(cell_group, parameter, c_effect, c_lower, c_upper, c_FDR)

# Pick the most significant one for demonstration
if(nrow(significant_cells) > 0) {
  selected_cell = significant_cells$cell_group[1]
  cat("\nSelected cell type for comparison:", selected_cell, "\n")
} else {
  # If no significant, pick one with largest effect
  selected_cell = test_results |>
    arrange(desc(abs(c_effect))) |>
    head(1) |>
    pull(cell_group)
  cat("\nNo significant cell types found. Using cell type with largest effect:", selected_cell, "\n")
}
```

### Simulate with Posterior Parameters (Real Effect)

```{r simulate-posterior, eval = instantiate::stan_cmdstan_exists(), message=FALSE}
# Simulate using all parameters from fit (includes real effects)
sim_posterior = sccomp_simulate(
  estimate,
  formula_composition = ~ type,
  formula_variability = ~ 1,
  new_data = NULL,
  coefficients = NULL,  # Use posterior beta_raw (real effects)
  
  number_of_draws = 10
)

# Calculate proportions for the selected cell type
# Use generated_proportions directly from the simulation output
proportions_posterior = sim_posterior |>
  filter(cell_group == selected_cell) |>
  select(sample, type, replicate, generated_proportions) |>
  rename(proportion = generated_proportions) |>
  mutate(scenario = "Posterior (real effects)")
```

### Simulate with Different Coefficient Scenarios

```{r simulate-scenarios, eval = instantiate::stan_cmdstan_exists(), message=FALSE}
# Create coefficients tables for different scenarios
coeffs_null = counts_obj |>
  distinct(cell_group) |>
  mutate(
    `(Intercept)` = 0,
    `typecancer` = 0  # No effect
  )

coeffs_positive = counts_obj |>
  distinct(cell_group) |>
  mutate(
    `(Intercept)` = 0,
    `typecancer` = if_else(cell_group == selected_cell, 1.5, 0.0)  # Positive effect for selected cell type
  ) |>
  mutate(`typecancer` = `typecancer` - mean(`typecancer`))  # Normalize to sum to zero

coeffs_negative = counts_obj |>
  distinct(cell_group) |>
  mutate(
    `(Intercept)` = 0,
    `typecancer` = if_else(cell_group == selected_cell, -1.5, 0.0)  # Negative effect for selected cell type
  ) |>
  mutate(`typecancer` = `typecancer` - mean(`typecancer`))  # Normalize to sum to zero

# Get sample data
sample_data = counts_obj |>
  distinct(sample, type)

# Simulate with different coefficients
sim_null = sccomp_simulate(
  estimate,
  formula_composition = ~ type,
  formula_variability = ~ 1,
  new_data = sample_data,
  coefficients = coeffs_null,  # Zero coefficients (no effect)
  
  number_of_draws = 10
)

sim_positive = sccomp_simulate(
  estimate,
  formula_composition = ~ type,
  formula_variability = ~ 1,
  new_data = sample_data,
  coefficients = coeffs_positive,  # Positive coefficients
  
  number_of_draws = 10
)

sim_negative = sccomp_simulate(
  estimate,
  formula_composition = ~ type,
  formula_variability = ~ 1,
  new_data = sample_data,
  coefficients = coeffs_negative,  # Negative coefficients
  
  number_of_draws = 10
)

# Calculate proportions for the selected cell type
# Use generated_proportions directly from the simulation output
proportions_null = sim_null |>
  filter(cell_group == selected_cell) |>
  select(sample, type, replicate, generated_proportions) |>
  rename(proportion = generated_proportions) |>
  mutate(scenario = "Zero (null)")

proportions_positive = sim_positive |>
  filter(cell_group == selected_cell) |>
  select(sample, type, replicate, generated_proportions) |>
  rename(proportion = generated_proportions) |>
  mutate(scenario = "Positive (+1.5)")

proportions_negative = sim_negative |>
  filter(cell_group == selected_cell) |>
  select(sample, type, replicate, generated_proportions) |>
  rename(proportion = generated_proportions) |>
  mutate(scenario = "Negative (-1.5)")
```

### Compare Proportions Between Scenarios

```{r compare-scenarios, eval = instantiate::stan_cmdstan_exists(), message=FALSE, fig.height=6}
# Combine all scenarios
comparison = bind_rows(
  proportions_posterior,
  proportions_null,
  proportions_positive,
  proportions_negative
)

# Calculate mean proportions by type and scenario
summary_comparison = comparison |>
  group_by(type, scenario) |>
  summarise(
    mean_proportion = mean(proportion, na.rm = TRUE),
    sd_proportion = sd(proportion, na.rm = TRUE),
    lower = quantile(proportion, 0.025, na.rm = TRUE),
    upper = quantile(proportion, 0.975, na.rm = TRUE),
    .groups = "drop"
  )

# Visualize comparison
p1 = comparison |>
  ggplot(aes(x = type, y = proportion, fill = scenario)) +
  geom_boxplot(position = "dodge", alpha = 0.7, outlier.alpha = 0.3) +
  facet_wrap(~scenario, ncol = 2) +
  theme_minimal() +
  labs(
    title = paste("Proportion Comparison:", selected_cell),
    subtitle = "Posterior (real) vs Zero vs Positive (+1.5) vs Negative (-1.5) coefficients",
    x = "Type", 
    y = "Proportion",
    fill = "Scenario"
  ) +
  theme(legend.position = "none")

# Show mean differences
p2 = summary_comparison |>
  ggplot(aes(x = type, y = mean_proportion, fill = scenario)) +
  geom_col(position = "dodge", alpha = 0.7) +
  geom_errorbar(
    aes(ymin = lower, ymax = upper),
    position = position_dodge(width = 0.9),
    width = 0.2
  ) +
  theme_minimal() +
  labs(
    title = "Mean Proportions with 95% CI",
    x = "Type",
    y = "Mean Proportion",
    fill = "Scenario"
  )

# Display plots
p1
p2

# Calculate difference between types for each scenario
# First, get mean proportion per type, scenario, and replicate
type_means = comparison |>
  group_by(type, scenario, replicate) |>
  summarise(mean_prop = mean(proportion, na.rm = TRUE), .groups = "drop")

# Get type levels to use correct column names after pivot
type_levels = sort(unique(as.character(type_means$type)))
type1 = type_levels[1]
type2 = type_levels[2]

# Calculate difference between types for each replicate
differences = type_means |>
  pivot_wider(names_from = type, values_from = mean_prop) |>
  mutate(
    difference = .data[[type2]] - .data[[type1]],
    abs_difference = abs(difference)
  )

# Show difference summary
cat("\nDifference between", type2, "and", type1, ":\n")
differences |>
  group_by(scenario) |>
  summarise(
    mean_diff = mean(difference, na.rm = TRUE),
    sd_diff = sd(difference, na.rm = TRUE),
    lower_diff = quantile(difference, 0.025, na.rm = TRUE),
    upper_diff = quantile(difference, 0.975, na.rm = TRUE),
    .groups = "drop"
  ) |>
  print()
```

This comparison clearly shows:
- **Posterior simulation**: Reflects the actual treatment effects estimated from the data
- **Zero coefficients simulation**: Shows what would happen if there were no treatment effects (null scenario)

The difference between these scenarios demonstrates the magnitude of the treatment effect for the selected cell type.

## 7. Summary

The `sccomp_simulate` function provides flexible data simulation capabilities:

- **Posterior Predictive Checks**: Use `new_data = NULL, coefficients = NULL` to simulate from the full posterior
- **New Conditions**: Provide `new_data` with new sample conditions while using posterior parameters
- **Custom Scenarios**: Provide `coefficients` table to specify custom effect sizes
- **Combined**: Use both `new_data` and `coefficients` for maximum flexibility
- **Effect Comparison**: Compare different coefficient scenarios to visualize treatment effects

All simulations respect the compositional nature of the data and include proper uncertainty quantification through the posterior distribution.

