---
title: "Smooth (spline) terms in sccomp"
author: "sccomp authors"
date: "`r Sys.Date()`"
package: sccomp
output:
  prettydoc::html_pretty:
    theme: cayman
    toc: yes
    toc_depth: 2
    number_sections: yes
    fig_caption: yes
    df_print: paged
abstract: >
  This vignette introduces `brms`-style smooth terms (`s()`, `t2()`) in
  `sccomp`. Smooths let you model non-linear effects of a continuous
  covariate — pseudotime, age, dose, time-since-treatment — on cell-type
  proportions, while keeping the rest of the `sccomp` modelling pipeline
  (compositional likelihood, variability modelling, outlier detection,
  random effects) unchanged. We start from the bundled `counts_obj`
  dataset, add a synthetic continuous covariate, fit a smooth, predict on
  a dense grid, and visualise both the fitted curves and the underlying
  basis decomposition.
vignette: >
  %\VignetteIndexEntry{Smooth (spline) terms in sccomp}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  fig.width  = 7,
  fig.height = 4.5,
  dpi = 110
)
# Smooth vignette chunks need mgcv (Suggests) and CmdStan.
run_smooth_vignette <-
  requireNamespace("mgcv", quietly = TRUE) &&
  instantiate::stan_cmdstan_exists()
```

# Why smooth terms?

`sccomp` fits a (logit-)linear model on the unconstrained scale and then maps
it to the simplex via softmax. If you only have categorical covariates this is
all you need. But with a **continuous covariate** — pseudotime, age, dose,
days since treatment — the relationship between covariate and proportion is
rarely linear. A spline (`s()` from `mgcv`) replaces a single linear column
with a small, flexible basis that lets the curve bend, while a penalty keeps
the bend honest.

`sccomp` adopts the same syntax as `brms` / `mgcv`:

```{r show-formulae, eval=FALSE}
~ type + s(pseudotime, k = 5)        # smooth of one variable
~ type + t2(pseudotime, dose)        # tensor product of two variables
```

Under the hood every smooth is split into two pieces:

* an unpenalised **null-space** block `Xf` (the linear "trend" part), and
* a penalised **wiggly** block `Xr` (the bend-y part).

`Xf` joins the design matrix as ordinary fixed-effect columns; `Xr` joins one
of `sccomp`'s random-effect slots and is governed by a single smoothing
parameter `sds`. This is exactly how `brms` handles smooths, and it means the
Stan model does **not** change — only the R-side data assembly.

# What is supported

| Feature                                            | Status                                                |
|----------------------------------------------------|-------------------------------------------------------|
| `s(x, k = ...)` in `formula_composition`           | yes (1 slot)                                          |
| `s(x, z, bs = "fs", k = ...)` (factor smooth)      | yes (3 slots: 1 wiggly + 2 per-level deviation blocks)|
| Multiple smooth terms in the same formula          | yes (each penalty block consumes one RE slot)         |
| Mix smooths with parametric terms (`~ type + s(x)`)| yes                                                   |
| Mix smooths with explicit `(... | g)` random effects | yes (total ≤ 4 slots)                              |
| Smooths in `formula_variability`                   | not yet                                               |
| `t2(x, y, ...)` (tensor-product smooth)            | fits, but `sccomp_predict()` not yet supported        |
| `s(x, by = factor)` (independent smooths per level) | works but consumes one slot **per level**            |

The 4-slot budget comes from the existing `sccomp` Stan model. If your
formula needs more than 4 random-effect slots in total (explicit `( | g)`
clauses + smooth penalty blocks), `sccomp_estimate()` will stop with an
informative error. Note that multi-penalty smooth types like `bs = "fs"`
emit several blocks; in mgcv / brms these correspond to several `sds_*`
parameters.

# Worked example

```{r load-libs, message=FALSE, warning=FALSE}
library(dplyr)
library(tidyr)
library(ggplot2)
library(sccomp)
data("counts_obj")
```

## 1. Synthetic continuous covariate

`counts_obj` has no continuous covariate, so we attach one — a *pseudotime*
between 0 and 6 — that varies across samples but is constant within a
sample (a covariate must always be sample-level in `sccomp`).

```{r make-pseudotime}
set.seed(1)
samples <- levels(counts_obj$sample)
pt_map <- tibble(
  sample     = samples,
  pseudotime = seq(0, 6, length.out = length(samples)) +
                 rnorm(length(samples), 0, 0.05)
)

counts_pt <- counts_obj |> left_join(pt_map, by = "sample")
counts_pt |> distinct(sample, type, pseudotime) |> head()
```

## 2. Fit `~ type + s(pseudotime, k = 5)`

```{r fit-smooth, message=FALSE, warning=FALSE, eval = run_smooth_vignette}
fit <- counts_pt |>
  sccomp_estimate(
    formula_composition = ~ type + s(pseudotime, k = 5),
    formula_variability = ~ 1,
    sample              = "sample",
    cell_group          = "cell_group",
    abundance           = "count",
    cores               = 1,
    verbose             = FALSE,
    max_sampling_iterations = 2000
  )
```

Look at the design-matrix column names that `sccomp_estimate()` prints. You
should see one new fixed column

```
s(pseudotime, k = 5)__lin1
```

and three penalised basis columns living in an RE slot

```
s(pseudotime, k = 5)___basis01
s(pseudotime, k = 5)___basis02
s(pseudotime, k = 5)___basis03
```

The `__lin*` columns are the null-space block (`Xf`); the `___basis*` columns
are the wiggly block (`Xr`). The single smoothing parameter `sds` for this
smooth controls how much the `___basis*` coefficients are pulled toward
zero — exactly the role of `sds_*` in `brms`.

## 3. Tabular output

`sccomp_test()` works unchanged. It returns one row per (cell_group,
parameter), and smooth basis columns appear like any other coefficient:

```{r test-table, eval = run_smooth_vignette}
fit |>
  sccomp_test() |>
  select(cell_group, parameter, c_lower, c_effect, c_upper, c_FDR) |>
  head(10)
```

`c_effect` on the `___basis*` rows is rarely interpreted directly — the
basis is internal — but the linear `__lin1` row is a clean "is there an
overall monotone trend in pseudotime for this cell group?" test, and the
basis-level `c_FDR` values let you spot non-linear effects.

## 4. Predict on a dense grid

To draw the fitted curve, predict at a regular grid of `pseudotime` values
(remember to also supply any other covariates the formula needs — here,
`type`):

```{r predict-grid, message=FALSE, warning=FALSE, eval = run_smooth_vignette}
grid <- expand.grid(
  pseudotime = seq(min(pt_map$pseudotime), max(pt_map$pseudotime), length.out = 60),
  type       = levels(counts_obj$type)
) |>
  as_tibble() |>
  mutate(sample = sprintf("grid_%03d", row_number()))

pred <- fit |>
  sccomp_predict(new_data = grid, number_of_draws = 200)

head(pred)
```

`sccomp_predict()` already returns the `new_data` columns (`pseudotime`,
`type`, `sample`) alongside `proportion_mean` and the 95% credible interval
`proportion_lower` / `proportion_upper`.

## 5. Plot: fitted curves over the raw proportions

We show the six cell groups with the largest pseudotime range in the
posterior mean.

```{r plot-curves, fig.width=8, fig.height=5, eval = run_smooth_vignette}
top_cgs <- pred |>
  group_by(cell_group) |>
  summarise(rng = max(proportion_mean) - min(proportion_mean)) |>
  arrange(desc(rng)) |>
  slice_head(n = 6) |>
  pull(cell_group)

raw <- counts_pt |>
  group_by(sample) |>
  mutate(prop = count / sum(count)) |>
  ungroup() |>
  filter(cell_group %in% top_cgs)

pred |>
  filter(cell_group %in% top_cgs) |>
  ggplot(aes(pseudotime, proportion_mean, colour = type, fill = type)) +
  geom_ribbon(aes(ymin = proportion_lower, ymax = proportion_upper),
              alpha = 0.15, linewidth = 0) +
  geom_line(linewidth = 0.7) +
  geom_point(data = raw, aes(y = prop), alpha = 0.5, size = 1.2) +
  facet_wrap(~ cell_group, scales = "free_y", ncol = 3) +
  labs(x = "pseudotime", y = "proportion",
       title = "Fitted compositional curves vs pseudotime") +
  theme_minimal(base_size = 10)
```

The blue/red curves are posterior means, ribbons are 95% credible intervals,
points are observed sample proportions. Note the wiggle — e.g. CD8 1 peaks
around pseudotime ≈ 4 — which a purely linear `~ type + pseudotime` model
could not capture.

## 6. Same model with a smaller basis (`k = 3`)

The same parametric + smooth layout works with a lower `k`. Fewer basis
functions mean a smoother curve and less risk of over-fitting when sample
size is modest — here `k = 3` yields one unpenalised linear column and a
single wiggly basis column (instead of three with `k = 5`).

```{r fit-smooth-k3, message=FALSE, warning=FALSE, eval = run_smooth_vignette}
fit_k3 <- counts_pt |>
  sccomp_estimate(
    formula_composition = ~ type + s(pseudotime, k = 3),
    formula_variability = ~ 1,
    sample              = "sample",
    cell_group          = "cell_group",
    abundance           = "count",
    cores               = 1,
    verbose             = FALSE,
    max_sampling_iterations = 2000
  )
```

The printed design matrix should list `(Intercept)`, `typecancer`, one
`__lin` column, and a single `___basis01` column for the smooth.

```{r predict-grid-k3, message=FALSE, warning=FALSE, eval = run_smooth_vignette}
pred_k3 <- fit_k3 |>
  sccomp_predict(new_data = grid, number_of_draws = 200)
```

```{r plot-curves-k3, fig.width=8, fig.height=5, eval = run_smooth_vignette}
pred_k3 |>
  filter(cell_group %in% top_cgs) |>
  ggplot(aes(pseudotime, proportion_mean, colour = type, fill = type)) +
  geom_ribbon(aes(ymin = proportion_lower, ymax = proportion_upper),
              alpha = 0.15, linewidth = 0) +
  geom_line(linewidth = 0.7) +
  geom_point(data = raw, aes(y = prop, colour = type), alpha = 0.5, size = 1.2) +
  facet_wrap(~ cell_group, scales = "free_y", ncol = 3) +
  labs(x = "pseudotime", y = "proportion",
       title = "Fitted curves: ~ type + s(pseudotime, k = 3)",
       subtitle = "One wiggly basis column — smoother than k = 5") +
  theme_minimal(base_size = 10)
```

Compared with the `k = 5` figure above, the credible bands are often a
little tighter and the lines less wiggly: the model has less capacity to
bend, which is appropriate when you want a gentle pseudotime trend on top
of the `type` effect.

# What `s()` actually decomposes into

The smooth metadata used to fit the model is stored on the result, so we
can re-evaluate the basis at any new pseudotime value. This is exactly what
`sccomp_predict()` does internally; we can also do it by hand to *see* the
basis.

```{r basis-decomp, fig.width=8, fig.height=6, eval = run_smooth_vignette}
smooth_specs   <- attr(fit, "smooth_specs")
smooth_re_objs <- attr(fit, "smooth_re_objs")

grid_basis <- tibble(
  pseudotime = seq(min(pt_map$pseudotime), max(pt_map$pseudotime),
                   length.out = 200),
  type = levels(counts_obj$type)[1]  # any constant works; smooth ignores it
)

basis <- sccomp:::predict_smooth_at_newdata(
  smooth_specs[[1]], smooth_re_objs[[1]], grid_basis
)
# Plain s() has one penalised block — for multi-penalty smooths
# (e.g. bs = "fs", t2(...)) `Xr_blocks` would contain several matrices.
Xr_single <- basis$Xr_blocks[[1]]

basis_long <- bind_rows(
  as.data.frame(basis$Xf) |>
    setNames(sprintf("lin%d", seq_len(ncol(basis$Xf)))) |>
    mutate(pseudotime = grid_basis$pseudotime) |>
    pivot_longer(-pseudotime, names_to = "column", values_to = "value") |>
    mutate(block = "Xf  (unpenalised, joins fixed effects)"),
  as.data.frame(Xr_single) |>
    setNames(sprintf("basis%d", seq_len(ncol(Xr_single)))) |>
    mutate(pseudotime = grid_basis$pseudotime) |>
    pivot_longer(-pseudotime, names_to = "column", values_to = "value") |>
    mutate(block = "Xr  (penalised, joins one RE slot, shared sds)")
)

ggplot(basis_long, aes(pseudotime, value, colour = column)) +
  geom_line(linewidth = 0.7) +
  facet_wrap(~ block, scales = "free_y", ncol = 1) +
  labs(x = "pseudotime", y = "basis value",
       title = "Decomposition of s(pseudotime, k = 5)",
       subtitle = "Linear null-space (Xf) is a fixed effect; wiggly basis (Xr) is one RE slot") +
  theme_minimal(base_size = 10)
```

The top panel is the **null-space**: for a default thin-plate `s()` this is
just the linear term in `pseudotime`. The bottom panel is the **wiggly
basis**: each of these columns gets one coefficient per cell group, and a
*single* smoothing standard deviation `sds` per cell group controls how
much wiggle is allowed. Increasing `k` would add more wiggly basis columns;
the penalty would then keep the extra capacity from over-fitting.

# Factor smooths: per-group curves with shared structure

A common ask is: "fit a smooth per tissue (or per condition, per donor, ...)
but partial-pool across groups". The `mgcv` shorthand for this is
`s(x, factor, bs = "fs")` — a *factor smooth*. It builds one common
wiggly curve and a per-level deviation, governed jointly by a small set of
smoothing parameters (three, in mgcv's standard reparameterisation: the
common wiggly part and two per-level deviation blocks). In `sccomp` each
of these blocks occupies its own RE slot.

## A worked example

We extend `counts_obj` with a synthetic `tissue` factor (3 levels) on top of
the existing `pseudotime`:

```{r make-tissue, message=FALSE}
set.seed(11)
samples <- levels(counts_obj$sample)
covs <- tibble::tibble(
  sample     = samples,
  pseudotime = seq(0, 6, length.out = length(samples)) +
                 rnorm(length(samples), 0, 0.05),
  tissue     = factor(rep(c("blood", "lymph", "tumor"),
                          length.out = length(samples)))
)
counts_tissue <- counts_obj |> dplyr::left_join(covs, by = "sample")
counts_tissue |> dplyr::distinct(sample, type, pseudotime, tissue) |> head()
```

Fit `~ s(pseudotime, tissue, bs = "fs", k = 5)`:

```{r fs-fit, message=FALSE, warning=FALSE, eval = run_smooth_vignette}
fit_fs <- counts_tissue |>
  sccomp_estimate(
    formula_composition = ~ s(pseudotime, tissue, bs = "fs", k = 5),
    formula_variability = ~ 1,
    sample              = "sample",
    cell_group          = "cell_group",
    abundance           = "count",
    cores               = 1,
    verbose             = FALSE,
    max_sampling_iterations = 2000
  )

# This one smooth occupies 3 RE slots (one per mgcv penalty block);
# you can confirm via:
attr(fit_fs, "model_input")$ncol_X_random_eff
```

The three non-zero entries correspond to the wiggly main block (9 columns
for `k = 5` × 3 tissues) and two smaller per-level deviation blocks.

## Predict and plot per-tissue curves

`sccomp_predict()` takes a grid that crosses `pseudotime` with `tissue`:

```{r fs-predict, message=FALSE, warning=FALSE, eval = run_smooth_vignette}
grid_fs <- expand.grid(
  pseudotime = seq(min(covs$pseudotime), max(covs$pseudotime),
                   length.out = 60),
  tissue     = levels(covs$tissue)
) |>
  as_tibble() |>
  mutate(sample = sprintf("fs_grid_%03d", row_number()))

pred_fs <- fit_fs |>
  sccomp_predict(new_data = grid_fs, number_of_draws = 200)

head(pred_fs)
```

Pick the most pseudotime-responsive cell groups and plot one curve per tissue:

```{r fs-plot, fig.width=8.5, fig.height=5.5, eval = run_smooth_vignette}
top_cgs_fs <- pred_fs |>
  group_by(cell_group) |>
  summarise(rng = max(proportion_mean) - min(proportion_mean)) |>
  arrange(desc(rng)) |>
  slice_head(n = 6) |>
  pull(cell_group)

raw_fs <- counts_tissue |>
  group_by(sample) |>
  mutate(prop = count / sum(count)) |>
  ungroup() |>
  filter(cell_group %in% top_cgs_fs)

pred_fs |>
  filter(cell_group %in% top_cgs_fs) |>
  ggplot(aes(pseudotime, proportion_mean,
             colour = tissue, fill = tissue)) +
  geom_ribbon(aes(ymin = proportion_lower, ymax = proportion_upper),
              alpha = 0.15, linewidth = 0) +
  geom_line(linewidth = 0.7) +
  geom_point(data = raw_fs, aes(y = prop, colour = tissue),
             alpha = 0.5, size = 1.2) +
  facet_wrap(~ cell_group, scales = "free_y", ncol = 3) +
  labs(x = "pseudotime", y = "proportion",
       title = "Factor smooth s(pseudotime, tissue, bs = 'fs', k = 5)",
       subtitle = "Per-tissue curves sharing one factor-smooth structure") +
  theme_minimal(base_size = 10)
```

The three coloured curves are the same smooth applied to different tissues,
allowed to deviate at each pseudotime point. Where the deviations are small
(e.g. CD8 1) the curves track each other; where they are large
(e.g. CD8 2) one tissue diverges. Whether the deviations are "real" is
expressed by the credible-interval ribbons — overlap means the data
don't support a difference.

## fs vs `s(x, by = tissue)`

`s(x, by = tissue)` is an alternative spelling that *also* gives a per-tissue
curve, but it treats each tissue's smooth as independent: each level gets
its own smoothing parameter, and the curves can have totally unrelated
shape. With L tissue levels this consumes L sccomp slots (one per level).
Use `bs = "fs"` when you want a single shared smoothness scale and partial
pooling across levels; use `by =` when the tissues are clearly heterogeneous
and a shared smoothness assumption would be misleading.


# Multiple smooths and mixing with random effects

`sccomp_estimate()` is happy with several smooths, and you can also keep
explicit `(... | g)` random-effect clauses:

```{r multi-smooth-formula, eval=FALSE}
~ type +
  s(pseudotime, k = 5) +
  s(age,        k = 4) +
  (1 | donor)
```

Each `s()` / `t2()` term consumes one of the four available RE slots.
The remaining slots are shared with any explicit `( | g)` clauses, so the
total must stay ≤ 4. If you exceed it, `sccomp_estimate()` will stop with
a message naming the smooth and the slot budget.

# Caveats and current limitations

* **No smooths in `formula_variability`** yet. `sccomp_estimate()` errors
  out if you try.
* **`t2()` tensor smooths** fit fine but `sccomp_predict()` is not yet
  supported for them: `mgcv::PredictMat()` applies the `t2` constraint
  differently from `smoothCon()`, producing a basis that doesn't line up
  with the fitted coefficients. The predict call errors out cleanly. Use
  `s(x, z, bs = "fs")` (shown above) when you want a smooth indexed by a
  factor, or stack univariate `s(x) + s(y)` smooths for additive
  multivariate effects.
* **`s(x, by = factor)`** decomposes into one independent smooth per factor
  level, each costing a separate RE slot. With many levels you will run
  out of the 4-slot budget. For partial-pooling across factor levels use
  `bs = "fs"` (shown above) — it fits in 3 slots regardless of the number
  of levels.
* The smooth columns are part of `sccomp_test()`'s output. The basis-level
  `c_effect` / `c_FDR` are diagnostics, not biological effect sizes;
  interpret the *curve* (predicted proportions) rather than individual
  basis coefficients.

# Session info

```{r session-info, echo=FALSE}
sessionInfo()
```
